---
title: "Applications of Item Response Theory in R"
author:
  - "W. Jake Thompson, Ph.D."
date: "`r Sys.Date()`"
output: ratlas::slides_html
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

class: reg-slide
layout: true

---
class: reg-section
# Materials


---

# Where can I get the stuff?

- Slides: epsy922.wjakethompson.com

- RStudio Cloud:

- Github: github.com/wjakethompson/r-irt

---

class: reg-section
# Heading Slide 2

---

# Slide 2

.pull-left[
Left text 
- bullet 1

- bullet 2

- bullet 3<sup>1</sup>
]

.pull-right[
Right text 
- bullet 4

- bullet 5

- bullet 6
]

--

.footnote[
[1] footnote for bullet 3
]

---


# R Plots

```{r cars}
par(mar = c(4, 4, 1, .1))
plot(cars, pch = 19, col = 'darkgray', las = 1) #<<
```

---

# Tables

If you want to generate a table, make sure it is in the HTML format (instead of Markdown or other formats), e.g.,

```{r}
knitr::kable(head(iris), format = 'html')
```

---
layout: false 
class: final-slide

# Connect With Us!

`r fontawesome::fa("globe", fill = "#165C7D")` [ATLAS4Learning.org](https://atlas4learning.org)

`r fontawesome::fa("envelope", fill = "#165C7D")` [ATLAS-AAI@ku.edu](mailto:atlas-aai@ku.edu)

`r fontawesome::fa("twitter", fill = "#165C7D")` [@ATLAS4Learning](https://twitter.com/atlas4learning)

`r fontawesome::fa("github", fill = "#165C7D")` [@ATLAS-AAI](https://github.com/atlas-aai)
